Hypertension and microbiome

### 2019-01-02 rrbiome.Rmd
White space
Functional

```{r variables}
data("phfinrisk_metadatadesc", package="finriskmetagcommon")
names.dset <- filter(phfinrisk_metadatadesc,
                     Ignored.Covariate.in.Cross.Sectional.Analysis.Aaro20181115==0)
```

@@

```{r prepare maaslin, eval = FALSE}

pheno <- t(meta(sample_data(pseq_filtered)))
sample.ids <- colnames(pheno)
phfg.pseq.rel <- microbiome::transform(pseq_filtered, "compositional")
abud.rel <- abundances(phfg.pseq.rel)

rownames(abud.rel) <- sapply(rownames(abud.rel),
                             function (genus) gsub("(.*) \\(+(.+)\\)", '\\1.\\2', genus))

merged <- rbind(sample.ids, pheno, abud.rel)
write.table(t(merged), file="merged_phfinrisk_genus.tsv", sep='\t', 
            quote=FALSE, row.names=FALSE, col.names=TRUE)

variables <- rownames(pheno)
variables_wo_ctrls <- variables[!variables %in% c("BL_AGE","BMI","SEX")]
data <- file("maaslin_config.config" , open='wt')
pcl.contents <- c("Matrix: Metadata",
                  paste(c("Read_PCL_Rows: BL_AGE,BMI,SEX",variables_wo_ctrls), collapse=","),
                  "",
                  "Matrix: Adundance",
                  "Read_PCL_Rows: Archaea.Archaea-")
writeLines(pcl.contents, con=data)
close(data)
```

```{r run maaslin, eval = FALSE} 
Maaslin("merged_phfinrisk_genus.tsv",
        "maaslin",
        strInputConfig = "maaslin_config.config",
        strModelSelection="none",
        dSignificanceLevel = 1.0, 
        fAllvAll = TRUE,
        strForcedPredictors = c("BL_AGE", "BMI", "SEX"))

@@

```{r Reading maaslins output, eval = FALSE}

maas.results.df <- read.table("maaslin/merged_phfinrisk_genus.txt", header=TRUE, sep='\t')
maas.results.df[,'Covariate'] <- maas.results.df[,'Variable']
maas.results.df <- merge(maas.results.df,
                         names.dset[,c('Covariate','Category', 'Name')],
                         by=c('Covariate'))

maas.results.df[,'Maaslin.Q.value'] <- maas.results.df[,'Q.value']
maas.results.df[,'Qval'] <- p.adjust(maas.results.df[,'P.value'], method="BH")
maas.results.df[,'beta'] <- maas.results.df[,'Coefficient']
maas.results.df.signf <- dplyr::filter(maas.results.df, Qval < 0.05)

```{r scatter plots, eval = FALSE}

sysm.signf.features <- maas.results.df.signf %>%
    dplyr::filter(Covariate == "SYSTM") %>%
    pull(Feature) %>%
    as.character() %>%
    gsub("_", ".", .)
df.merged <- t(merged) %>% as.data.frame
```

```{r maaslin scatter plot, fig.width = 10, fig.height = 20, dpi = 300, eval = FALSE}

#df.merged %>% pull("Rikenella.Bacteria") %>% as.character I%>% as.numeric %>% sqrt %>% asin

ggplot(df.merged %>%
       select(sysm.signf.features, SYSTM, SEX) %>%
       gather(Covariate, Value, sysm.signf.features) %>%
       mutate(SYSTM = as.numeric(as.character(SYSTM)),
              Value = as.numeric(as.character(Value))),
       aes(x = asin(sqrt(Value)), y = SYSTM)) +
    facet_wrap(~Covariate, ncol=2, scales = "free") +
    geom_point(aes(color = SEX), size = 0.5) +
    scale_y_continuous(breaks=seq(0,200,40)) +
    geom_smooth(method=lm, se=TRUE, size=0.5, color = 'Black') +
    stat_poly_eq(formula = "y ~ x",
                 aes(label = paste(..eq.label.., ..rr.label.., sep = "~~~")),
                 label.y = 0,
                 parse = TRUE,
                 size = 3) +   
    ggplot2::theme_minimal() 
```

>>>>

```{r variables}
data("phfinrisk_metadatadesc", package="finriskmetagcommon")
names.dset <- filter(phfinrisk_metadatadesc,
                     Ignored.Covariate.in.Cross.Sectional.Analysis.Aaro20181115==0)
names.dset <- bind_rows(names.dset,
                        data.frame(Covariate = c("MAP", "PULSEPRESSURE", "HYPERTENSION"),
                                   Category = rep("Physical", 3),
                                   Name =c( "Mean artrerial pressure", "Pulse pressure", "Hypertension")))
```

@@

```{r prepare maaslin}

get.merged.dataset <- function(pseq) {
    pheno <- t(microbiome::meta(phyloseq::sample_data(pseq)))
    sample.ids <- colnames(pheno)
    phfg.pseq.rel <- microbiome::transform(pseq, "compositional")
    abud.rel <- abundances(phfg.pseq.rel)
    rownames(abud.rel) <- sapply(rownames(abud.rel), function (genus) gsub("(.*) \\(+(.+)\\)", '\\1.\\2', genus))
    t(rbind(sample.ids, pheno, abud.rel))
}

prepare.maaslin <- function (pseq,
                             tsvfile = "merged_phfinrisk_genus.tsv",
                             conffile = "maaslin_config.config",
                             includedvars = c("BL_AGE", "BMI", "SEX")) {
    write.table(get.merged.dataset(pseq), file = tsvfile, sep='\t', quote=FALSE, row.names=FALSE, col.names=TRUE)
    data <- file(conffile , open='wt')
    pcl.contents <- c("Matrix: Metadata",
                      paste(c("Read_PCL_Rows: ", paste(includedvars, collapse=",")), collapse = ""),
                      "",
                      "Matrix: Adundance",
                      "Read_PCL_Rows: Archaea.Archaea-")
    writeLines(pcl.contents, con=data)
    close(data)
}

 merge.pheno.abu <- function(pseq) {
     pheno <- microbiome::meta(phyloseq::sample_data(pseq))
     sample.ids <- rownames(pheno)
     phfg.pseq.rel <- microbiome::transform(pseq, "compositional")
     abud.rel <- abundances(phfg.pseq.rel)
     rownames(abud.rel) <- sapply(rownames(abud.rel), function (genus) gsub("(.*) \\(+(.+)\\)", '\\1.\\2', genus))
     cbind(pheno, t(abud.rel))
}
```

Running maaslin for blood pressure variables while rest of the
phenotype variables are forced in the model

```{r run maaslin} 
maaslin.force.vars <- c("BL_AGE", "SEX", "BMI", "CURR_SMOKE", "Q57X", "DIAB", "BL_USE_RX_C03", "BL_USE_RX_C07", "BL_USE_RX_C08", "BL_USE_RX_C09")

prepare.maaslin(pseq.genus,
                 tsvfile = "maaslin_merged_phfinrisk_genus.tsv",
                conffile = "maaslin_config.config",
                includedvars = c("MAP", "SYSTM", "DIASM", "PULSEPRESSURE", "HYPERTENSION", maaslin.force.vars))

Maaslin("maaslin_merged_phfinrisk_genus.tsv",
        "maaslin",
        strInputConfig = "maaslin_config.config",
        strModelSelection="none",
        dSignificanceLevel = 1.0, 
        fAllvAll = TRUE,
        strForcedPredictors = maaslin.force.vars)
```

```{r Reading maaslins output}
maas.results.df <- read.table("maaslin/merged_phfinrisk_genus.txt", header=TRUE, sep='\t')
maas.results.df[,'Covariate'] <- maas.results.df[,'Variable']
maas.results.df <- merge(maas.results.df, names.dset[,c('Covariate','Category', 'Name')], by=c('Covariate'))
maas.results.df[,'Maaslin.Q.value'] <- maas.results.df[,'Q.value']
maas.results.df[,'Qval'] <- p.adjust(maas.results.df[,'P.value'], method="BH")
maas.results.df[,'beta'] <- maas.results.df[,'Coefficient']
maas.results.df.signf <- dplyr::filter(maas.results.df, Qval < 0.05)

@@

```{r scatter plots}
sysm.signf.features <- maas.results.df.signf %>%
    pull(Feature) %>%
    as.character() %>%
    gsub("_", ".", .)

df.merged <-  merge.pheno.abu(pseq.genus)
```

```{r maaslin scatter plot, fig.width = 10, fig.height = 20, dpi = 300}
ggplot(df.merged %>% gather(Covariate, relabundance, sysm.signf.features), aes(x = asin(sqrt(relabundance)), y = SYSTM)) +
    facet_wrap(~Covariate, ncol=2, scales = "free") +
    geom_point(aes(color = SEX), size = 0.2) +
    scale_y_continuous(breaks=seq(0,200,40)) +
    geom_smooth(method=lm, se=TRUE, size=0.5, color = 'Black') +
    stat_poly_eq(formula = "y ~ x",
                 aes(label = paste(..eq.label.., ..rr.label.., sep = "~~~")),
                 label.y = 0,
                 parse = TRUE,
                 size = 3) +   
    ggplot2::theme_minimal() 
```

```{r maaslin plot}
source("edgeR_plot_functions.R")

plot_maaslin_heatmap_final(maas.results.df.signf,
                      xlab.text="Genera",
                      beta.maxabs=max(abs(maas.results.df.signf[,'beta'])),
                      order.by.category=FALSE)
```
